Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimization Choices

### 2025-10-27 revision/response.qmd

1. The paper introduces two interesting new developments to projection pursuit research: the quantification of index properties that enables better comparison, and a new optimization approach based on Jellyfish Search Optimizer. The results show the good performance of the new optimization strategy, and how it can be related to the new metrics introduced. While the overall presentation is clear, it feels like the authors are conflicted about what the main focus of the paper actually should be. The introduction mentions that the primary goal is to investigate the potential for the jellyfish search optimization, however the title does not even mention this, and much of the paper seems to focus on the new metrics introduced. Having a more well defined focus should help guide the flow of the paper better (currently it feels like jumping around between the two objectives).

**\textcolor{purple}{TODO: Di}**.

2. I also think additional clarifications are needed in the definition of squintability. It seems to me Definition 4 is simply defining an angle between planes (without reference to "squinting"), while in the following the definition of squintability is based on index values at given projection distances?

**\textcolor{purple}{Squint angle is the core concept inspires the metric squintability, hence we provide a separate definition of it in Definition 4. For all the projection bases in the space, those near the optimal basis correspond to projections with interesting structure we're reesarching. While for different indexes, the size of the neighbourhood is different: some indexes may have projections resemble the target even when it is far away, which corresponds to a large squint angle and easy to optimize (since it is easy for the optimizer to sample into the neighbourhood). To quantify how close projections need be before structures of interest can be observed (the optimizer starts hill-climbing), we can track how the index value changes as we move from an initial projection basis to the optimal one (projection distance). This prompts the definition of projection distance in Definition 5 and squintability in Definition 6. We have revise the text to explain this connection more clearly in Section 3.2.}**

@@

6. The results in section 6 are very interesting. However, I was wondering about some details in 6.3: from the results shown it seems that the smoothness and squintability for the selected data examples are highly correlated, could this cause issues with the model fit shown in Table 2? It would be interesting if a similar result for CRS could be included, to see the effect of smoothness in that setting.

**\textcolor{purple}{The correlation between the two variables are 0.55, which should not affect the modelling. (TODO: Di - do you mind checking if we need more here?) We did try CRS for the same sine-wave optimisation reported in Section 6, but most runs yield a failed result with the optimum not found by CRS, hence not included.}**

7. I was also wondering how the dependence of the smoothness on d can be explained, since it is not intuitive that this should change so much? (Note also that the last sentence in the caption for Table 1 seems to mix up squintability and smoothness.)

**\textcolor{purple}{Table 1 is a subset of the PP index estimated and it happens that the three chosen ones all decrease as n goes larger. This is not necessary the case in the full table shown in the Appendix. The caption for Table 1 has been fixed.}**

@@

8. Finally, I appreciate the practical advice in section 7. Here, 7.1 should be somewhat extended in my opinion: does the planned tour then run through all the individual paths, should we follow the most successful jellyfish, or how would we pick?

**\textcolor{purple}{We make Section 7.1 clearer that the planned tour can be run to trace the optimisation of individual jellyfish path. Users may wish to follow the most successful jellyfish, a random selection, or interesting jellyfish that result in local optimum. DISCUSSION: should I include a small example with the code?}**

Finally I have also collected a few minor remarks:

  * After defining Eq. 4 the additional parameter should probably be theta_4 and not theta_2.  **\textcolor{purple}{TODO: Nicolas}**
  * In the algorithm pseudo-code it is not clear what c_t and the trend actually are.  **\textcolor{purple}{In the text and the pseudo-code, we include information that $c_t$ is the time control function determine the exploration and exploitation phase of the algorithm. We also include information that the (ocean) trend is defined as the difference between the best jellyfish and the average of all current jellyfish.}**
  * In Fig. 4 please specify what optimizer was used. **\textcolor{purple}{We have included the information in the caption that the CRS optimised is used for this illustration. }**
  * From the last sentence of Section 5, it is not clear to me why the ranks need to be used, why not scale the metrics?  **\textcolor{purple}{We include further explanation in the end of Section 5 on the choice of using rank for the modelling. The reason is because of the numerical instability due to the small squintability value associated with the string index (less than 0.002). Scaling is less ideal here because we define the two metrics to be between [0, 1 ] and scaling will distort the interpretation of the GLM coefficients.}**.
  * In 6.2 it is not really clear what is being bootstrapped for the evaluation of the uncertainty. **\textcolor{purple}{We have included further description in the paper to describe the bootstrap procedure as drawing 500 bootstrap samples with replacement from the original 50 simulations for each setting: an optimisation with high-dimensional data (d = 4, 6, 8, 10, or 12) with different number of jellyfish (20, 50, or 100) and maximum number of iteration (50 or 100)}**.

### 2025-10-27 jso.qmd
more revision

From @barnett1981interpreting and @laa_using_2020, a large squint angle
implies that the objective function value is close to optimal even when
the target view to see the structure is far away. A small squint angle
means that the PP index value improves substantially only when the
target view is close by. As such, low squintability implies rapid
improvement in the index value when near the target view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimizer needs to be very close to be able to "see" the optimum, making it difficult for the optimizer to find the optimum.

@@

It is expected that for a PP index with high squint angle, the
optimization \eqref{eq-optimization} should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation, shown in the simulated trace from different index functions in @fig-squintability-ill, suggests that the half-point index value improvement percentage provides an appropriate
mathematical definition of squintability, which matches the intuition
behind this concept, while being amenable to numerical computation.

@@

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ be a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $r(A, A^*)$ to the index value $f(XA)$, such that $g(r) = g(r(A, A^*)) = f(XA)$.  The squintability of an index function $f$ is defined by

@@

A generalized linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting the success rate.
Predictors are smoothness, squintability, data dimension, and JSO
hyper-parameters. Due to the scale of the smoothness and squintability measure,
the rank of the two metrics, rather than the nominal values, are used in the model.

>>>>

The squintability metric is inspired by the concept of squint angle in Definition \ref{def:squint-angle}, originally introduced in @barnett1981interpreting and later discussed in @laa_using_2020. Consider the space of all projection bases: those near the optimal basis correspond to projections with interesting structure we're researching. The size of this neighborhood depends on the choice of the index function. For some indexes, projections resemble the target even when relatively far from the optimum, corresponding to a large squint angle, while for others, projections need to be close to the optimum to reveal structure we're interested, resulting in a small squint angle. For PP, a small squint angle is considered to be undesirable because it means that the optimizer needs to sample into a be very narrow region to be able to "see" the optimum, making it difficult for the optimizer to find the optimum.

@@

To quantify how close projections need be before structures of interest can be observed (the optimizer starts hill-climbing), we can track how the index value changes as we move from an initial projection basis towards the optimal one, that is, how index value changes as the projection distance becomes smaller.

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ be a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}

@@

@fig-squintability-ill illustrates some simulated traces for different PP index functions: MIC, splines2d, skinny, and stringy2. It is expected that for a PP index with high squint angle, the optimization \eqref{eq-optimization} should make substantial progress early on. Conversely, for a PP index with low squint angle, it might take a long while for the optimization to make substantial progress, as the candidate projections would need to be very close to the optimal one for the structure of the index function to be visible enough to be amenable to efficient optimization. This observation suggests that the half-point index value improvement percentage provides an appropriate
mathematical definition of squintability, which matches the intuition
behind this concept, while being amenable to numerical computation.

\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $r(A, A^*)$ to the index value $f(XA)$, such that $g(r) = g(r(A, A^*)) = f(XA)$.  The squintability of an index function $f$ is defined by

@@

A generalized linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting the success rate.
Predictors are smoothness, squintability, data dimension, and JSO
hyper-parameters. For numerical stability related to small squintability values associated with the stringy index (squintability less than 0.002), we decide to use the rank of the two metrics, rather than the nominal values, in the model.

### 2025-10-27 jso.tex
more revision

From Tukey and Tukey (1981) and Laa and Cook (2020), a large squint
angle implies that the objective function value is close to optimal even
when the target view to see the structure is far away. A small squint
angle means that the PP index value improves substantially only when the
target view is close by. As such, low squintability implies rapid
improvement in the index value when near the target view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimizer needs to be very close to be able to ``see'' the optimum,
making it difficult for the optimizer to find the optimum.

@@

It is expected that for a PP index with high squint angle, the
optimization \eqref{eq-optimization} should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation, shown in the
simulated trace from different index functions in
Figure~\ref{fig-squintability-ill}, suggests that the half-point index
value improvement percentage provides an appropriate mathematical
definition of squintability, which matches the intuition behind this
concept, while being amenable to numerical computation.

@@

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ be a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $r(A, A^*)$ to the index value $f(XA)$, such that $g(r) = g(r(A, A^*)) = f(XA)$.  The squintability of an index function $f$ is defined by

@@

\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, opacityback=0, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, colback=white, leftrule=.75mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, arc=.35mm, coltitle=black, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm]

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in
Figure~\ref{fig-proportion}. Bootstrap resampling is used to quantify
the uncertainty of the success rate by drawing 500 bootstrap samples
with replacement from the original 50 simulation results for each task:
optimizing a high-dimensional data (d = 4, 6, 8, 10, or 12) with
different number of jellyfish ( 20, 50, or 100) and maximum number of
iteration (50 or 100). As the number of jellyfish and maximum iteration
allowed increase, the success rate also increases. For problems in lower
dimensional (4, 6) search spaces, small parameter values (20 jellyfish
and 50 iterations) are sufficient to achieve a high success rate. Many
more jellyfish (100 jellyfish and 100 iterations) are needed for finding
the pipe structure in d = 10, and 12. This suggests that even JSO will
struggle with high dimensions, and require substantially more
computational time to have an acceptable success rate, especially for
difficult PP indexes based on scagnostics.

>>>>

The squintability metric is inspired by the concept of squint angle in
Definition \ref{def:squint-angle}, originally introduced in Tukey and
Tukey (1981) and later discussed in Laa and Cook (2020). Consider the
space of all projection bases: those near the optimal basis correspond
to projections with interesting structure we're researching. The size of
this neighborhood depends on the choice of the index function. For some
indexes, projections resemble the target even when relatively far from
the optimum, corresponding to a large squint angle, while for others,
projections need to be close to the optimum to reveal structure we're
interested, resulting in a small squint angle. For PP, a small squint
angle is considered to be undesirable because it means that the
optimizer needs to sample into a be very narrow region to be able to
``see'' the optimum, making it difficult for the optimizer to find the
optimum.

@@

To quantify how close projections need be before structures of interest
can be observed (the optimizer starts hill-climbing), we can track how
the index value changes as we move from an initial projection basis
towards the optimal one, that is, how index value changes as the
projection distance becomes smaller.

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ be a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}

@@

Figure~\ref{fig-squintability-ill} illustrates some simulated traces for
different PP index functions: MIC, splines2d, skinny, and stringy2. It
is expected that for a PP index with high squint angle, the optimization
\eqref{eq-optimization} should make substantial progress early on.
Conversely, for a PP index with low squint angle, it might take a long
while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the half-point
index value improvement percentage provides an appropriate mathematical
definition of squintability, which matches the intuition behind this
concept, while being amenable to numerical computation.

\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $r(A, A^*)$ to the index value $f(XA)$, such that $g(r) = g(r(A, A^*)) = f(XA)$.  The squintability of an index function $f$ is defined by

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, toprule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, breakable, colback=white, leftrule=.75mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, bottomrule=.15mm, coltitle=black, bottomtitle=1mm, opacitybacktitle=0.6, left=2mm, opacityback=0]

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of iteration) on the success rate is summarized in
Figure~\ref{fig-proportion}. Bootstrap resampling is used to quantify
the uncertainty of the success rate through 500 bootstrap samples with
replacement from the original 50 simulation results for each task:
optimizing a high-dimensional data (d = 4, 6, 8, 10, or 12) with
different number of jellyfish ( 20, 50, or 100) and maximum number of
iteration (50 or 100). As the number of jellyfish and maximum iteration
allowed increase, the success rate also increases. For problems in lower
dimensional (4, 6) search spaces, small parameter values (20 jellyfish
and 50 iterations) are sufficient to achieve a high success rate. Many
more jellyfish (100 jellyfish and 100 iterations) are needed for finding
the pipe structure in d = 10, and 12. This suggests that even JSO will
struggle with high dimensions, and require substantially more
computational time to have an acceptable success rate, especially for
difficult PP indexes based on scagnostics.
